\documentclass[fignum,nobf,doc,noapacite]{apa}
\newtheorem{lemma}{Lemma}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\R}{R}
\def\citeapos#1{\citeauthor{#1}'s (\citeyear{#1})}

\usepackage{eurosym}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amsbsy}
\usepackage{listings}
\usepackage{setspace}
\usepackage{pdfcomment}
\usepackage{subcaption}
\usepackage{pdfpages}		% to include the full-page plot
\usepackage{authblk}		% for mulitple authors with different affiliations

\usepackage[american]{babel}
\usepackage{csquotes}
\usepackage[style=apa,natbib=true, backend=biber]{biblatex}
\DeclareLanguageMapping{american}{american-apa}


\usepackage[T1]{fontenc}
\usepackage[sc]{mathpazo}
\usepackage{color}
\linespread{1.05}         % Palatino needs more leading (space between lines)

\bibliography{bibfile}

\abstract{Experiments suggest that there is a strong tendency to verbally recode visually-presented information, and that in some cases verbal recoding can boost memory performance.  According to multi-component models of working memory, memory performance is increased because task-relevant information is maintained in two codes. The possibility of dual encoding is problematic if the goal is to measure capacity for visual information exclusively. This is why articulatory suppression is typically used with visual change detection tasks specifically to prevent verbalization of visual stimuli. We present evidence from a large experiment that suggests that articulatory suppression has no discernable effect on performance in a visual change-detection task. Neither a descriptive nor a state-trace analysis revealed any complex relationship between suppression, presentation type, and performance. There is no evidence that verbal recoding was a strategy used by the participants in this task. We conclude that in typical visual change detection experiments, pre-cautionary articulatory suppression is unnecessary.}


\rightheader{}
\shorttitle{}

\leftheader{}


\title{Opportunity for verbalization does not improve visual change detection performance: A state-trace analysis.}

\author[1]{Florian Sense}
\author[2]{Candice C. Morey}
\author[3]{Melissa Prince}
\author[4]{Andrew Heathcote}
\author[5]{Richard D. Morey}

\affil[1]{Department of Experimental Psychology \& Department of Psychometrics and Statistics, University of Groningen}
\affil[2]{Department of Psychology, University of Edinburgh}
\affil[3]{School of Psychology, Flinders University, Australia}
\affil[4]{School of Medicine, The University of Tasmania, Australia}
\affil[5]{School of Psychology, Cardiff University}

\lstset{tabsize=2}
\acknowledgements{Please address correspondence to: {\tt f.sense@rug.nl}. The code for all analyses and plots as well as the complete dataset can be found at {\tt www.linktosupplement.com}.

We would like to thank Yongqui Cong, Christian Hummeluhr, and Mareike Kirsch for valuable assistance with data collection.}


\begin{document}
\maketitle


\newpage

\section{Introduction}
\input{intro.tex}


\section{Methods}
\input{methods.tex}


\section{Results}
\input{results.tex}


\section{Discussion and Conclusions}
\input{discussion.tex}


\printbibliography

\end{document}
